\documentclass[11pt]{article}

% ---------------------------------------------------------------------------
% The Neuroarchitectural Transformer — arXiv preprint (also Gradient Paper XI).
% Compile: pdflatex main && bibtex main && pdflatex main && pdflatex main
% (or: latexmk -pdf main.tex). No non-standard .sty files required.
% ---------------------------------------------------------------------------

\usepackage[margin=1in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{array}
\usepackage{microtype}
\usepackage[numbers,sort&compress]{natbib}
\usepackage[hidelinks]{hyperref}
\usepackage{xcolor}
\usepackage{enumitem}

\newcommand{\zone}[1]{\textsc{#1}}
\newcommand{\status}[1]{\textit{[#1]}}
\newcommand{\Qff}{Q16.16}

\title{\textbf{The Neuroarchitectural Transformer:}\\[2pt]
Verifiable-by-Construction, Capability-per-Parameter,\\ and Paraconsistent Federated Training}

\author{Larry Klosowski\\ Mozi Satori / Citrate Network\\ \texttt{@saulbuilds}
\and The Claude Opus 4.x research-and-implementation lineage}

\date{June 2026 \quad (RFC-CIT-NAT-0001 / Gradient Paper XI)}

\begin{document}
\maketitle

\begin{abstract}
Modern language models route every parameter through every computation, so there is no honest
answer to the question \emph{``which part of the model produced this?''} That opacity is tolerable
for casual chat and disqualifying for regulated, on-chain, or scientific use, where a decision must
be replayable and a contribution must be verifiable. We present the \textbf{Neuroarchitectural
Transformer (NAT)}, a transformer whose hidden representation is partitioned into a fixed set of
\textbf{declared, named zones}---each with its own core (attention or state-space)---that execute in
parallel, communicate over a \textbf{fixed topology} whose edge strengths a learned router only
\emph{modulates} (it cannot create undeclared edges), and whose outputs are combined by an
attention-scored, noise-pruned \textbf{merge running on a deterministic \Qff{} fixed-point path}.
Every forward pass emits a structured, deterministically-hashable \textbf{provenance trace} of which
zones fired, with what confidence, what was pruned, and why; the trace is the product, making the
model interpretable and verifiable by construction, on every pass, rather than after the fact. We
argue this single architectural move---\emph{declaring structure}---yields three properties usually
pursued separately. \textbf{Verifiability} is demonstrated: the trace is a committable, replayable
artifact whose decision layer is third-party-checkable, and the stateful surfaces are model-checked
in TLA+. \textbf{Capability per parameter} is empirically supported at small scale: on a 1.12M-token
public-domain corpus trained as a next-byte language model with a held-out split, an equal-parameter
ablation finds that zone partitioning \emph{does not reduce} capability per parameter---and is
modestly lower-loss on the mean---across five seeds (NAT held-out loss 2.88--2.91 vs.\ dense
2.97--2.99; a non-inferiority verdict, $N{=}5$, no formal significance test), with a learned router
that generalizes to held-out prompts (separation 3.10 vs.\ 2.63). \textbf{Decentralizability} is
specified and formally modeled, not yet demonstrated: composable named zones reconciled via
paraconsistent (Belnap) aggregation on a verifiable chain. All empirical results are at small scale
($\sim$20K--115K parameters, byte-level), without a mixture-of-experts baseline or component
ablation, and we are explicit that a larger, better-controlled run could refute them. NAT is
implemented as a reproducible Rust workspace; a GGUF/ONNX sidecar design targets ecosystem
compatibility. We position NAT as the model layer for a verifiable, federated learning network in
which \emph{consensus and learning are the same process}, and contributors are rewarded as
$\mathrm{compute}\times\mathrm{data\text{-}quality}$.
\end{abstract}

% ===========================================================================
\section{Introduction}

\paragraph{The blob of weights.}
A transformer \citep{vaswani2017attention} is, mechanically, an undifferentiated block of parameters
through which every token is pushed in full. This is enormously effective and enormously opaque. When such a model emits an
output, there is no architectural fact of the matter about \emph{which part of it} was responsible;
any account of ``what the model was doing'' must be reconstructed, after training, by external
tooling (probing, sparse autoencoders, activation patching). Mechanistic interpretability has made
real progress at this reconstruction \citep{rai2024practical,bereska2024mechanistic}, but it remains
\emph{post-hoc}: it approximates, after the fact, a structure the architecture never committed to.
You cannot \emph{prove} what ran; you can only \emph{infer} it. For chat that is acceptable; for
regulated decisions, on-chain inference where a decision must be replayable, and decentralized
scientific training where a contribution must be verifiable to be rewarded, opacity is
disqualifying. These settings do not want a better explanation of a black box. They want the box not
to be black.

\paragraph{The thesis: structure is interpretability.}
NAT's wager is that one can largely dissolve the opacity problem by \emph{declaring} structure the
architecture must respect, rather than discovering it afterward. We \textbf{partition} the hidden
representation into named zones, each mapped---as a \emph{mimetic analog}, not a fidelity claim---to
a functional role; \textbf{wire} them over a fixed, declared topology that an auditor and a model
checker can both read, modulated per input by a learned router that cannot create an undeclared
edge; \textbf{merge} the outputs by attention-scored pruning on a deterministic \Qff{} fixed-point
path; and \textbf{emit} a structured provenance trace as a first-class output of every pass. The
trace is not a debug aside; it is the deliverable.

\paragraph{One move, three properties.}
The contribution we most want to land is that declaring structure is not a trade.
\textbf{(1) Verifiable by construction.} Zero-knowledge ML can prove, expensively and after the
fact, that \emph{some} opaque computation ran on committed weights
\citep{chen2024zkml,sun2024zkllm}---verifying the output, not the reasoning. NAT's trace is
verifiable by construction: the recorded decision is decision-faithful and replayable with no
per-inference SNARK, on the same \Qff{} substrate as Citrate's verifiable-inference precompiles
\citep{citrate10}. \textbf{(2) Capability per parameter that does not pay an interpretability tax.}
Holding parameters, data, seed, and compute fixed and varying only the partitioning, zone
partitioning does not reduce capability per parameter versus an equal-parameter dense baseline---and
is modestly lower-loss on the mean---across all five seeds we ran (a non-inferiority result, small
scale, no MoE baseline or component ablation yet; \S\ref{sec:results}).
\textbf{(3) Decentralizable.} Because zones are composable, a federation can train and evolve a
single zone; independently-trained contributions reconcile through the deterministic merge locally
and, across nodes, through paraconsistent (Belnap) aggregation \citep{belnap1977useful} that
preserves disagreement (\emph{Both}) and ignorance (\emph{Neither}) rather than averaging them away,
operationalizing the Citrate thesis that consensus and learning are one process \citep{citrate2}.

\paragraph{Contributions.}
We tag each by status: \status{demonstrated} (implemented and evaluated), \status{implemented}
(built and tested, not an evaluation), \status{specified} (designed, where noted formally modeled,
not yet demonstrated).
\begin{enumerate}[leftmargin=2em,itemsep=1pt]
\item \status{implemented} The \textbf{zone-partitioned architecture}: declared named zones over a
fixed, auditable topology with learned-but-bounded routing (the router provably cannot create
undeclared edges), hybrid SSM/attention cores by function, and a non-learned executive harness.
\item \status{demonstrated} \textbf{Provenance-as-verifiable-output}---the primary
contribution: a first-class, deterministically-hashable trace, decision-faithful and
third-party-replayable on the \Qff{} path and on-chain-committable, distinct from model cards,
logging, and post-hoc interpretability, with a precise decision-faithful vs.\ bit-faithful
distinction.
\item \status{specified} An \textbf{ecosystem-compatibility design}: a GGUF/ONNX sidecar; the
flattened-dense export and Ollama-class round-trip are not yet built.
\item \status{demonstrated, small scale} The \textbf{H-01 result}: an equal-parameter ablation on
real text finding partitioning does not reduce capability per parameter (non-inferiority, modest
mean advantage, 5 seeds)---with explicit scale and missing-baseline caveats.
\item \status{specified} A \textbf{paraconsistent federated-training frame}: composable zones
reconciled by Belnap aggregation on a verifiable \Qff{} substrate (local determinism TLA+-checked;
the multi-node cycle is future work), with $\mathrm{compute}\times\mathrm{data\text{-}quality}$
incentives.
\item \status{implemented} A \textbf{reproducible Rust reference implementation} with TLC-green
TLA+ specifications of the stateful surfaces, and a companion case study on agent-led model building.
\end{enumerate}

\paragraph{Honest posture.}
Every quantitative claim is anchored to a file or commit, or labeled a hypothesis; the brain analogy
is a heuristic the architecture is free to abandon; and the load-bearing capability claim is stated
with its scale caveat in the same breath. If a larger run refutes H-01, the honest move is to say so
and change course---and the scale ladder exists so we, and the reader, can find out cheaply.

% ===========================================================================
\section{Related work}\label{sec:related}

\paragraph{Modular deep learning and mixture-of-experts.}
Modular architectures route inputs through a subset of parameter-efficient modules and aggregate the
result \citep{pfeiffer2023modular}; sparse mixture-of-experts realizes this for transformers
\citep{shazeer2017outrageously,lepikhin2020gshard,fedus2022switch,jiang2024mixtral}. NAT shares the
conditional-computation instinct but inverts the routing's epistemics: MoE experts are discovered,
interchangeable, and unnamed, and the gate is a black box. We do not claim named modules or fixed
routing are new---hash/fixed-routed MoE \citep{roller2021hash} fixes the assignment, named modules
are surveyed in \citet{pfeiffer2023modular}, and named, fixed-topology functional modules are decades
old in cognitive architectures (ACT-R, Leabra). What is distinctive is not the naming but that the
topology is a \emph{machine-checkable auditable object}---a learned router that \emph{provably} cannot
create an undeclared edge (\S\ref{sec:arch}, with the merge and gather model-checked,
\S\ref{sec:verif})---composed with the provenance trace. MoE optimizes throughput; NAT optimizes
auditability and capability-per-parameter at equal total parameters.

\paragraph{Brain-inspired language models.}
The closest prior art is BriLLM \citep{zhao2025brillm}, which also takes neuroscience as its starting
point and claims full interpretability via a ``signal fully-connected flowing'' (SiFu) mechanism over
a graph that \emph{replaces attention entirely}. The differences are decisive: BriLLM abandons the
transformer and with it the GGUF/ONNX/Ollama ecosystem; NAT keeps the transformer and stays
ecosystem-compatible through a sidecar (\S\ref{sec:arch}). BriLLM's interpretability is node-level
token mapping; NAT's is functional-zone provenance recorded as a deterministically hashable,
on-chain-verifiable artifact, not merely a readable graph (\S\ref{sec:verif}). And NAT advances a
falsifiable per-parameter capability claim against an equal-parameter dense baseline, which BriLLM
does not.

\paragraph{State-space and hybrid sequence models.}
State-space models \citep{gu2021s4,gu2023mamba,dao2024mamba2} offer linear-time recurrence and an
explicit state; hybrids such as Jamba \citep{lieber2024jamba} interleave SSM and attention
\emph{layers}. NAT organizes the hybrid by declared function: SSM cores in the temporal zones (\zone{SM},
\zone{CB}), attention in the reasoning zones (\zone{HP}, \zone{PF}, \zone{CX}), each SSM zone carrying a
thin attention head for cross-zone talk.

\paragraph{Interpretability.}
Mechanistic interpretability reverse-engineers trained transformers
\citep{rai2024practical,bereska2024mechanistic} and is overwhelmingly \emph{post-hoc}: it
reconstructs, after training, a structure the architecture never committed to, and so approximates
but cannot prove what ran. NAT makes interpretability intrinsic by construction---the zones are
declared, and the provenance trace makes ``which functional component produced this'' a recorded,
replayable fact.

\paragraph{Verifiable and zero-knowledge ML.}
ZkML proves, after the fact, that an opaque computation ran on committed weights
\citep{chen2024zkml,sun2024zkllm,qu2025zkgpt,peng2025survey}, at heavy cost---zkLLM takes on the order
of fifteen minutes per proof for a 13B model, and even zkGPT, the fast path, proves GPT-2 in tens of
seconds---and says nothing about the reasoning. NAT is verifiable by construction: the
decision-faithful trace is replayable with no per-inference proof, on the same \Qff{} substrate as
Citrate's verifiable-inference precompiles \citep{citrate10}, with which it composes when bit-exact
certification of the numeric layer is required (\S\ref{sec:verif}).

\paragraph{Decentralized and federated training.}
Federated averaging \citep{mcmahan2017communication} and low-communication distributed training
\citep{douillard2023diloco,prime2024opendiloco} train \emph{monolithic} models as periodically
synchronized replicas. NAT federates composable named zones instead of whole replicas: a node owns
and trains a zone, submits signed zone outputs, and the deterministic merge reconciles
independently-trained contributions (\S\ref{sec:fed}). Where DiLoCo averages, NAT aggregates with
paraconsistent semantics.

\paragraph{Decentralized science and blockchain AI.}
Networks such as Bittensor and Gensyn, and the broader DeSci/DeAI movement \citep{wang2024sok},
incentivize \emph{opaque} contributions scored by validators---you trust the score, not the
computation. NAT supplies the verifiable substrate that line lacks: a contributor's work is a signed,
provenance-traced, \Qff{}-deterministic update whose value is
$\mathrm{compute}\times\mathrm{data\text{-}quality}$ (\S\ref{sec:fed}).

\paragraph{Paraconsistent logic.}
Belnap--Dunn's four-valued logic \citep{belnap1977useful}, with values true, false, both, and
neither, is designed for reasoning from multiple inconsistent or incomplete sources; its continuing
imprint in computer science is traced in \citet{jakl2025four}. NAT uses it as the
federated-aggregation logic (\S\ref{sec:fed}), preserving genuine disagreement and ignorance as
first-class.

\paragraph{Scaling laws.}
Compute-optimal scaling \citep{hoffmann2022training,kaplan2020scaling} holds architecture fixed and
trades off parameters, data, and compute. H-01 asks the orthogonal question those laws hold constant:
at fixed parameters, data, seed, and compute, does \emph{structure} change capability per parameter?

% ===========================================================================
\section{The NAT architecture}\label{sec:arch}

\subsection{Zones}
NAT partitions a transformer's hidden representation of width $D$ into a fixed set of named zones,
each owning a contiguous slice, its own sequence core, input projection, and internal state. There
are six zones; five are learned and one is a non-learned executive harness (Table~\ref{tab:zones}).
The mapping to brain regions is a mimetic analog, not a fidelity claim; the load-bearing facts are
that the zones are declared, the partition is fixed at build time, and the assignment of a
representational slice to a named function is a property of the architecture rather than a discovery
made after training. All six zones are declared and the forward pass is exercised end-to-end on all
six (the L0 reference), but the \emph{capability} evidence in \S\ref{sec:results} is on the three
data-rich zones $\{\zone{HP},\zone{PF},\zone{CX}\}$: the multimodal \zone{SM} and timing \zone{CB} zones
have the thinnest data at this scale, so we hold them out of the load-bearing ablation until ``the
data earns it'' (a staging decision). The architecture is six zones; the measured claims are three.

\begin{table}[t]\centering\small
\begin{tabular}{@{}lll@{}}
\toprule
Zone & Role & Core \\
\midrule
\zone{SM} Sensorimotor & ingest + temporally bind multimodal input & state-space (SSM) \\
\zone{CB} Cerebellar & timing, sequencing, learned reflex & state-space (SSM) \\
\zone{HP} Hippocampal & memory consolidation, novelty/salience & attention \\
\zone{PF} Prefrontal & reasoning, planning, language (deepest) & attention \\
\zone{CX} Codec & reasoning $\rightarrow$ verifiable executable logic & attention \\
\zone{MX} MCP harness & validate / sequence / route tool use & none (state machine) \\
\bottomrule
\end{tabular}
\caption{The six zones. Temporal zones use state-space cores for linear-time recurrence and an
explicit, loggable state; reasoning zones use attention. The SSM recurrence is computed as a single
lower-triangular matmul, device-agnostic across CPU and GPU. The core backend is pluggable and
recorded in every trace.}
\label{tab:zones}
\end{table}

\subsection{Routing: fixed topology, learned modulation}
The permitted inter-zone edges are declared in the sidecar and fixed at build time (default:
$\zone{SM}{\to}\{\zone{CB},\zone{HP},\zone{PF}\}$, $\zone{CB}{\to}\zone{PF}$, $\zone{HP}{\to}\zone{PF}$,
$\zone{PF}{\to}\zone{CX}$). For each input the router produces a zone-activation vector
$a\in[0,1]^{Z}$ and an edge-modulation weight $m_e\in[0,1]$ for each \emph{declared} edge. The crucial
property---the one that keeps the system auditable while adaptive---is that the router can only
modulate declared edges; \textbf{it cannot create an edge that is not there.} In the implementation
this is structural: the router iterates the declared edge set and nothing else, so an undeclared edge
has no code path to receive a weight. This is claim-shaped for IP review (C-1).

\subsection{Parallel execution and async gather}
Zones execute in parallel and finish at different times. The merge boundary uses an async gather:
each zone returns its output tagged with a confidence and a status; the merger waits up to a deadline,
then composes with whatever arrived, recording a late zone as \texttt{timed\_out}. This is the same
gather discipline the federated case needs (\S\ref{sec:fed}); its safety (no zone both \texttt{ok}
and \texttt{timed\_out}) and liveness (the window always closes) are model-checked
(\S\ref{sec:verif}).

\subsection{The merge}
The merge runs four ordered steps on the gathered outputs: \textbf{score} each output; \textbf{prune}
the bottom 70--80\% by score (recording pruned contributions with their scores, so the decision is
auditable not silent); \textbf{re-weight} the survivors to sum to one; \textbf{compose} by weighted
sum. Steps 2--3 are a single canonical function used both to \emph{produce} the trace and to
\emph{verify} it, so the verification in \S\ref{sec:verif} is not circular. The compose step runs on
the deterministic \Qff{} fixed-point path: a value $v$ is the integer $\mathrm{round}(v\cdot 2^{16})$
in an \texttt{i64}, with \texttt{i128} multiply intermediates. Integer arithmetic is bit-identical
across CPUs and federated nodes where IEEE-754 float is not---the property that lets independently
computed merges reconcile and on-chain provenance verify. The training-time merge is a differentiable
form \emph{reconciled to} this \Qff{} provenance merge: a battery test pins that its hardened survivor
set equals the canonical survivors (the soft top-$k$ via temperature-annealed softmax converges to
the hard survivors as $\tau\to 0$), so the model the gradient sees and the model the verifier replays
make the same survivor decision; the composition weights are soft during training and anneal toward
the hard recorded weights.

\subsection{The MCP harness}
Tool use is mediated by a non-learned executive harness (\zone{MX}): a deterministic state machine plus
a validator. It walks its states in order ($\textsc{input\_validation}\to\cdots\to\textsc{return}$)
with two safety guards that fail closed: \textbf{no external side effect occurs before the action gate
approves}, and \textbf{a failed Codec verification can never reach tool execution}. Because it is
non-learned and side-effect-ordered, it is the most straightforwardly formalizable component; the
TLA+ module states exactly these two invariants plus termination (claim C-4).

\subsection{Serialization: the GGUF sidecar}
NAT keeps GGUF/ONNX as the tensor container and adds a sidecar (\texttt{.nat.json}) declaring the zone
graph, topology, and merge parameters. A sidecar-unaware runtime runs the tensors opaquely; a
sidecar-aware runtime runs the full zone pass with provenance. We are precise about status, because it
is not yet built: a literal zone-partitioned graph with parallel heterogeneous SSM+attention zones
does not serialize to a layout \texttt{llama.cpp} runs as-is, so the ecosystem onramp depends on a
\emph{flattened-dense} export that is \textbf{specified, not yet implemented}; the sidecar declares an
\texttt{export\_kind} field whose \texttt{FlattenedDense} variant is reserved for it, but only the
\texttt{ZonePartitioned} form is produced today. The claim that NAT ``runs in the existing inference
ecosystem'' is a design target, not a measured result.

% ===========================================================================
\section{Verifiability by construction}\label{sec:verif}

\subsection{The trace}
Every forward pass emits a structured provenance trace: an \texttt{input\_hash}, the core
\texttt{backend} (\texttt{toy-l0} / \texttt{candle-cpu} / \texttt{candle-cuda}), the router's
modulation, per-zone records (id, core, activated, confidence, latency, status), inter-zone flows,
the merge (scores, prune threshold, survivors, weights), the codec verification, the harness
transitions and tool calls, and an \texttt{output\_hash}. It must be \emph{complete} (every activated
zone and prune decision recorded), \emph{hashable} (deterministic serialization---stable field order,
fixed-point values as raw integers, no map iteration), and \emph{faithful} (next).

It is worth saying what is \emph{not} novel: ``logging what the model did'' is old. Model cards
\citep{mitchell2019model}, datasheets, and audit logs document a model or a run; activation caching
records internals for post-hoc analysis. The trace differs on three specific counts, and the novelty
is their conjunction: it is a \textbf{first-class output of the forward pass itself} (not an external
probe); it is \textbf{decision-faithful and deterministically replayable} by a third party on the
integer \Qff{} path, so the routing/prune decision can be re-derived bit-exactly rather than merely
inspected; and it is \textbf{committed on-chain} as part of the inference transaction. No model-card
or logging scheme offers third-party deterministic replay of the decision.

\subsection{Decision-faithful vs.\ bit-faithful}
A naive reading of ``replaying the logged zone mix reproduces the output'' cannot be bit-exact,
because the learned cores run in floating point and float is not deterministic across hardware. We
distinguish two levels and claim only what holds. \textbf{Decision-faithful}: replaying the recorded
scores through the canonical merge decision reproduces the recorded survivors and weights---a pure
integer computation that always holds, and is what the verifier checks; because the same function
produces and verifies the decision, this is non-circular rather than tautological in the sense that it
confirms the implementation matches the specification. \textbf{Bit-faithful}: re-running the full pass
reproduces \texttt{output\_hash} bit-for-bit, which holds only under a deterministic-inference path
(the \Qff{} merge is deterministic; the float cores are deterministic only in a deterministic mode).
The decision-faithful guarantee is claim C-2; bit-faithful is an optional mode, and it composes with
the cryptographic layer (\S\ref{sec:verif}.4).

\subsection{On-chain commitment and replay}
On Citrate, the trace hash becomes part of the inference transaction; an auditor pulls the committed
weights, replays the recorded decision, and confirms it reproduces the output. That replayability is
the opacity solution: we record the computation in a form a third party can re-check.

\subsection{Relation to Citrate's verifiable-inference substrate}
NAT does not replace verifiable inference; it supplies the structural layer zkML and TEE attestation
lack. ZkML proves, after the fact and at heavy cost, that some opaque computation ran on committed
weights, verifying the output not the reasoning. NAT's trace verifies the reasoning, by construction,
with no per-inference proof, because its merge uses the same \Qff{} substrate as Citrate's
verifiable-inference precompiles \citep{citrate10}: a NAT trace hash is directly committable, and when
bit-exact certification of the numeric layer is needed it can be wrapped in a Halo2-KZG proof or a
TEE attestation as that work describes.

\subsection{Formal specifications}
Three TLA+ modules \citep{lamport2002specifying} cover the stateful surfaces: \texttt{MergeDeterminism} (same gathered set
$\Rightarrow$ same result; pruning correct, complete, tie-free), \texttt{AsyncGather} (terminates,
respects the deadline, records every straggler), and \texttt{McpHarness}
(\textsc{NoUngatedSideEffect}, \textsc{NoExecOnFailedCodec}, termination). \textbf{TLC is green on all
three} (via \texttt{scripts/run-tlc.sh}): \texttt{MergeDeterminism} over 31 distinct states,
\texttt{AsyncGather} over 40, \texttt{McpHarness} over 10, with no invariant violations. The remaining
open formal item is counsel sign-off on the claim-shaped statements C-1--C-5, not the model checking.
The defensible combination is \emph{auditable-by-construction, GGUF-compatible, zone-partitioned, with
on-chain-verifiable provenance and a formally-specified tool harness.}

% ===========================================================================
\section{Hypotheses and experimental design}\label{sec:hyp}

\paragraph{The ledger.}
\textbf{H-01} (load-bearing): zone partitioning does not reduce capability per parameter versus a
dense baseline of equal size. \textbf{H-02}: context-aware routing produces measurably different zone
mixes for different prompt classes. \textbf{H-03a/b}: provenance is decision-faithful / bit-faithful
(under a deterministic path). \textbf{H-04}: SSM temporal zones cut per-zone compute. \textbf{H-05a/b}:
the merge composes the same gathered set identically / a federated training cycle reproduces the
centralized result within tolerance (a distinct, statistical claim). H-01 decides whether the bet
pays off; the scale ladder tests it cheaply.

\paragraph{The H-01 protocol.}
An ablation at unequal parameters proves nothing, so the comparison is pinned by protocol and the
protocol is \emph{enforced in code}---the harness refuses to run (and to report) a comparison outside
tolerance. The partitioned arm is the real trainable model (zones + the learned router + the
differentiable merge reconciled to the \Qff{} provenance merge); the control is an equal-parameter
dense single-block transformer whose feed-forward width is searched until its parameter count matches
within $\pm5\%$ (the achieved match here is 0.08\%: 20{,}718 vs.\ 20{,}701); a second guard refuses a
toy-backed arm, so a measured result cannot be a toy artifact. Both arms share the same data, windows,
epochs, batch size, learning rate, and shuffle seed---only the structure differs. Capability is
proxied as the inverse of held-out cross-entropy, measured per parameter, averaged over five seeds;
since $1/\mathrm{loss}$ is an arbitrary monotone transform and the arms are matched to 0.08\%, the
comparison is in effect \emph{lower held-out loss at equal parameters}. The per-seed verdict is the
non-inferiority test ``partitioned $\geq$ dense within a 5\% slack,'' and the headline is the
holds-fraction. We ran H-01 first on a synthetic task (too smooth, as it turned out) and then on real
text, which we treat as primary.

\paragraph{The corpus.}
The L1 corpus is a 1.12M-token, license-clean, public-domain ``values spine'' built through a
fail-closed pipeline (license, length, dedup, PII, and quality gates). Its curation thesis doubles as
a thesis of the paper: \emph{a rule has no meaning without a community and a form of life}
(Wittgenstein)---which is why a maker follows the rules of the room, why provenance must answer to a
public standard, and why good code reads like the codebase around it. It spans logic (Boole
$\to$ Frege $\to$ Russell $\to$ Belnap), computation (Turing $\to$ Church $\to$ Shannon), craft
(SICP, the Rust Book, permissive code), and expression (Strunk, Whitman, Montaigne). The pipeline
reports 2{,}337 documents, 779 shards, 1{,}120{,}711 tokens, and a self-scored aggregate quality of
0.852 (a curation diagnostic, not an external measure). Tokenization is byte-level (vocabulary 256),
so the metric is bits/byte (uniform baseline 8.0).

% ===========================================================================
\section{Results}\label{sec:results}

\subsection{H-01: partitioning does not reduce capability per parameter}
Under the protocol of \S\ref{sec:hyp}, the real model and an equal-parameter dense transformer
(20{,}718 $\approx$ 20{,}701 parameters) were each mini-batch-trained on the 1.12M-token corpus as a
next-byte LM, capability measured as the inverse of held-out cross-entropy per parameter across five
seeds. \textbf{H-01 is supported: partitioning does not reduce capability per parameter, and is
modestly lower-loss on the mean, across all five seeds.} We state this carefully: H-01 is registered
as a \emph{non-inferiority} hypothesis, and the per-seed verdict is a non-inferiority test with a 5\%
slack, so ``5/5'' means ``on no seed did partitioning fall more than 5\% behind.'' With that defined,
the partitioned model's held-out next-byte loss is \textbf{2.88--2.91 versus the dense baseline's
2.97--2.99} (per-seed extremes, non-overlapping), with mean capability-per-parameter $1.670\times
10^{-5}$ vs.\ $1.621\times10^{-5}$. The ranges are non-overlapping and consistent with a real effect
in NAT's favor, but $N{=}5$ with no within-arm variance reported and no formal significance test, so
we call the result \emph{suggestive and supportive}, not a demonstration of superiority.

A synthetic pre-read (binned-token-sum, $\approx$3{,}882 params) held on the mean but on 3 of 5
seeds; 3/5 vs.\ 5/5 at $N{=}5$ is not statistically distinguishable, the synthetic and real runs
differ in task, batching, and metric, and we treat the real-data run as primary because it is the
more realistic setting, not because 5/5 is decisive where 3/5 was not. The scale caveat is
$\sim$20K parameters, three zones, byte-level, $\sim$1M tokens; a parameter-matched mixture-of-experts
control and component ablations (\S\ref{sec:disc}) are the experiments that would establish
\emph{which} structural feature carries the effect.

\subsection{H-02: a trained router differentiates and generalizes}
The trained router drives measurably different zone mixes for different prompt classes and
\emph{generalizes rather than memorizes}: on a held-out split of the evaluation battery (math /
narrative / code / sensory), it separates classes it never saw more sharply than the unlearned
baseline, \textbf{3.10 versus 2.63} on the held-out separation metric. On the in-sample battery the
gap is larger (11.70 vs.\ 4.25), which is the optimistic upper read; we treat the held-out 3.10/2.63
as the honest evidence. Full-scale labeled batteries across more classes are the conclusive read.

\subsection{A size/zone ladder trends downward}
\begin{table}[h]\centering\small
\begin{tabular}{@{}lrrr@{}}
\toprule
Rung & Params & Zones & Held-out bits/byte \\
\midrule
S & 20{,}718 & 3 & 4.097 \\
M & 56{,}534 & 3 & 4.054 \\
L & 114{,}956 & \textbf{5} & \textbf{3.953} \\
\bottomrule
\end{tabular}
\caption{A three-rung ladder; held-out bits/byte fall at each step. The ladder is confounded: the L
rung changes parameters \emph{and} zone count (3$\to$5), so we cannot attribute its improvement to
size alone. Three points are a trend, not a scaling law.}
\label{tab:ladder}
\end{table}
Holding the corpus fixed, held-out bits/byte fall at each step of a three-rung ladder
(Table~\ref{tab:ladder}). Two cautions keep this from being a scaling claim: three points are a
trend, not a law; and the L rung changes \emph{two} variables at once (parameters and zone count
3$\to$5), so we cannot attribute its improvement to size alone. The ladder is suggestive evidence the
architecture does not \emph{degrade} with size at this range, not proof that it scales. Separately, a
per-position autoregressive objective reached 3.42 bits/byte at 53K parameters; we note this only as
a denser objective the path to L2 will use---it is \emph{not} a controlled size or efficiency
comparison against the single-output ladder, and we do not present it as one.

\subsection{Decision-faithfulness}
Decision-faithfulness is a \textbf{design property, not an empirical result}: replaying the recorded
scores reproduces the recorded survivors and weights because the same function produces and verifies
the decision, so the check confirms the implementation matches the specification and cannot, by
construction, disconfirm the property. Bit-faithfulness is empirical and mode-dependent: it holds at
the deterministic L0 scale and only under a deterministic-inference mode at L1.

\subsection{Summary}
The load-bearing hypothesis is supported (non-inferiority, modest mean advantage) at L1 on real data
across five seeds, with honest small-scale and no-significance-test caveats; routing differentiation
generalizes at small scale (held-out 3.10 vs.\ 2.63); the size/zone ladder trends downward over three
confounded rungs; and the verifiability guarantee holds by construction with the stateful surfaces
TLC-checked. What we have \emph{not} shown---a parameter-matched MoE baseline and component ablations,
a task-level metric, a standard corpus, the hold surviving orders-of-magnitude more scale, and
federated training in practice---is the subject of \S\ref{sec:disc} and \S\ref{sec:concl}.

% ===========================================================================
\section{Federated training by paraconsistent consensus}\label{sec:fed}

The third face of the thesis is that declaring structure makes the model decentralizable. Much of this
layer is \emph{specified}, not \emph{demonstrated}, and we mark the boundary throughout.

\paragraph{Composable zones.}
Because a zone owns a slice of fixed width and a declared cross-zone contract, it is swappable when
its slice width and contract match---so a federation can train and evolve a single zone without
retraining the whole model. A node owns one or more zones, trains them, and submits its zone outputs;
the async gather \emph{would} collect them under the deadline discipline used inside a single pass,
and the deterministic \Qff{} merge \emph{would} reconcile independently-computed contributions such
that every node and an on-chain verifier compute identically. We use the conditional deliberately: the
determinism property (same gathered set $\Rightarrow$ same bits) is demonstrated locally and
TLC-checked, and the gather is implemented today as a single-process deterministic simulation of the
deadline; the multi-node, cross-network signed gather is the federated milestone.

\paragraph{Paraconsistent aggregation.}
Federated nodes have different data distributions, and averaging discards the signal that
distinguishes a healthy decentralized network from a collapsed one. NAT adopts the mechanism specified
in \citet{citrate2}: aggregate per dimension in Belnap's four-valued logic \citep{belnap1977useful}.
Each dimension resolves to a state---\emph{True}, \emph{False}, \emph{Both} (sources disagree:
contradictory), or \emph{Neither} (no source spoke: unknown)---and only T/F collapse to a mean over
consenting sources; a \emph{Both} dimension is routed to multiple zones for cross-validation rather
than averaged, and a \emph{Neither} dimension is left undefined rather than fabricated. The
aggregation runs over \Qff{} embeddings, so it is bit-deterministic and on-chain-checkable. This
operationalizes the Citrate thesis that consensus and learning are the same process at different time
scales: the same checkpoint that finalizes blocks finalizes the Belnap-aggregated zone weights.
\emph{Status:} the aggregation logic and its on-chain commitment are specified and supported by
deployed learning-cycle contracts \citep{citrate2}; the end-to-end federated training cycle is future
work.

\paragraph{The incentive seam.}
NAT does not implement settlement; it emits a signed contribution---metered compute, a data-quality
score, a token count, and the provenance-trace hash---and a deterministic reward weight $=
\mathrm{compute}\times\mathrm{data\text{-}quality}$, computed on the \Qff{} path so two nodes and an
on-chain verifier agree bit for bit. \texttt{citrate-compute-pool} converts that weight to payout. A
node that contributes compute on garbage data earns weight zero, which is why the pipeline's quality
stage is load-bearing.

\paragraph{Decentralized intelligence meets decentralized science.}
Decentralized-AI networks today incentivize opaque contributions scored by trusted validators; NAT's
provenance trace makes every training and inference step verifiable by construction, on a substrate
the chain already runs. A contributor's work is not a black box a validator must trust; it is a
signed, replayable, provenance-traced artifact anyone can check---a model not served \emph{to} a
community but \emph{trained by} one, whose every step answers to a public standard.

% ===========================================================================
\section{Discussion}\label{sec:disc}

\paragraph{Why one move yields three properties.}
Once the partition is declared and the merge is deterministic, the same fact---which named function
contributed, with what weight---is simultaneously a verifiable record, an inductive bias that
empirically does not hurt at equal parameters, and a unit of contribution a federation can sign and
reconcile. We did not design three mechanisms; we declared one structure and read three properties
off it.

\paragraph{Threats to validity.}
\textbf{Scale.} Every quantitative result is at $\sim$20K--115K parameters, byte-level, $\sim$1M
tokens; a larger model, BPE, more depth, and more data could change the verdict; a confounded
three-rung curve is not a scaling law. \textbf{Missing baselines and ablations}---the most important
gap. H-01 compares NAT against exactly one control, an equal-parameter dense transformer; there is
\emph{no parameter-matched mixture-of-experts baseline}, the obvious comparison for a paper positioned
against MoE, and \emph{no component ablation} isolating which feature carries the effect (router vs.\
the pruning merge vs.\ the partition vs.\ SSM/attention heterogeneity). The $\sim$0.08--0.10 loss gap
could be a regularization effect of the pruning merge or of using several narrower sub-blocks; our
experiments cannot rule that out. \textbf{Non-standard corpus.} Results are on a bespoke corpus, not
a standard one (enwik8/text8/WikiText), so the numbers are non-comparable to external results.
\textbf{Statistics.} Five seeds, no within-arm variance, no significance test; 3/5 and 5/5 are not
distinguishable at $N{=}5$. \textbf{Metric.} The capability proxy is inverse held-out loss, not a
task-level metric. \textbf{Specified vs.\ demonstrated.} The federated/Belnap layer is specified;
bit-faithfulness is mode-dependent. \textbf{The brain analogy is not yet earned.} We have not shown
that the neuro-motivated zone assignment outperforms a \emph{random} equal-width partition; until that
ablation is run, the neuroscience framing is motivation, not evidence. \textbf{Single operator.} One
machine, self-reported, no external replication; the reproducibility floor exists to make replication
cheap.

\paragraph{Honest posture as method, and an agent-built model.}
The discipline that produced this paper is itself a claim: every capability claim points at a
measurement or is labeled a hypothesis, and the marginal synthetic H-01 read was reported as marginal,
which is why the work went and got real data rather than declaring victory. The architecture was
designed in a human--AI conversation; the reference implementation, the data pipeline, the training
stack, and the ablation were built largely by AI agents under that discipline; and the paper was
drafted by an agent and then attacked by an adversarial multi-agent red-team that found over- and
under-claims and forced each line back to what the repository supports. A companion case study
documents this. The point is not novelty but evidence: a verifiable architecture and a logged,
reproducible method answer to the same standard---the same discipline that makes the model's
provenance trustworthy makes the research record trustworthy.

% ===========================================================================
\section{Conclusion and future work}\label{sec:concl}

We presented the Neuroarchitectural Transformer, a transformer partitioned into declared, named zones
over a fixed, auditable topology, merged on a deterministic fixed-point path, emitting a hashable
provenance trace as a first-class output. The wager---that declaring structure dissolves much of the
opacity problem and pays out in verifiability, capability per parameter, and decentralizability at
once---is supported, at small scale, by what we measured: on a real public-domain corpus, partitioning
did not reduce capability per parameter and was modestly lower-loss on the mean across all five seeds
(a non-inferiority result); a learned router generalized to held-out prompts; held-out loss trended
downward over a three-rung ladder; and decision-faithful provenance holds by construction with the
stateful surfaces TLC-checked. These are small-scale results without a mixture-of-experts baseline or
component ablation, stated with their caveats.

The most important next experiments are the ones that would turn the H-01 result from suggestive to
causal: a parameter-matched mixture-of-experts baseline; component ablations (no-router, no-prune,
single-core-type, and a \emph{random} equal-width partition versus the named one, to earn the
neuroscience framing); and results on a standard corpus with a per-seed table, variance, and a paired
significance test, ideally with one independent replication. Scale is the next test after that; the
federated proof turns \S\ref{sec:fed} from specified to demonstrated; the GGUF flattened-dense export
builds the ecosystem onramp the sidecar currently only specifies; and a task-level metric replaces the
inverse-loss proxy. Each is a falsification opportunity as much as a milestone. If the broader claim
holds---that a model can be a dynamic, legible, verifiable instrument rather than an undifferentiated
blob of weights, with a verifiable record on every pass---then the contribution is not only an
architecture but a stance: that interpretability, efficiency, and decentralization are not three
problems but three views of one decision, \emph{declare the structure}.

\paragraph{Reproducibility.}
The reference implementation is a commit-pinned Rust workspace with config hashes, fixed seeds, exact
rerun commands, and a containerized CI path; the three TLA+ modules are TLC-green; the H-01 run and
the corpus build are reproducible from single commands. Code and data-pipeline are available in the
\texttt{nat} repository.

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
